Filling in data gaps on refugee children: Statistical imputation for missing count data

### 2023-01-24 paper.qmd
Remove file

---
title: "Filling in data gaps on refugee children: Statistical imputation for missing count data"
author: Jan Beise, Yukun Pei (UNICEF); Sebastian Steinmueller (UNHCR)
abstract: UNHCR's annually published population dataset with the number of refugees and their age/sex composition by country of origin and country of asylum contained missing age data for 23 per cent of the global refugee populations at end-2021. Figures of the global proportion of refugees who are children disseminated to date used the available data only. We show that this approach likely leads to upwards biased estimates of the global proportion of children due to the omission of refugee groups in the Americas and Europe with lower proportions of children. We propose a multilevel Bayesian model to create a dataset with multiple imputations of age/sex counts for country pairs with missing data, resulting in standardisation of global and regional estimates by countries of origin and regions of asylum. The estimate of the proportion of children in the global refugee population derived from the imputed dataset is at 42 per cent compared to 47 per cent in the previously communicated available-data figure. 
format: pdf
bibliography: references.bib
execute:
  echo: false
  warning: false
  message: false
header-includes:
    \usepackage{float}
    \usepackage{setspace}
    \onehalfspacing
toc: true
lof: true
lot: true
---

@@

# Demographic data and data gaps on refugees

@@

@@

While the imputed dataset is useful in estimating regional and global estimates of the age/sex distribution of refugees, the modelling approach is top-down and uses only geographic proximity and regional groupings as a means to improve available-data figures. It cannot substitute data collected and compiled at the national level. If the main goal is to obtain reliable estimates of the demographic characteristics of refugees on the level of country of asylum, strengthening the production process of official statistics on refugees in said country should be the primary course of action. Recommendations and implementation examples developed by countries in the framework of the Expert Group on Refugee, IDP and Statelessness Statistics (see EGRISS 2018) can provide helpful guidance to national statistical systems. Projecting populations from census, survey or administrative data with some data and reasonable assumption on fertility, mortality and displacement are another way to obtain more reliable estimates of the age/sex distribution at country level, although they require much more effort for a single country and are therefore better suited if estimates for a specific country are required. Estimates from the imputed dataset at any geographical level should always be disseminated and communicated with uncertainty intervals or by showing the posterior predictive distribution over imputations. While care has been taken to include and test prior assumptions and hierarchical geographic structures to model uncertainty from parameters and predictions, we made a blanket assumption that all available age/sex data was measured without uncertainty and without measurement error. This is an over-simplification. For example, counts for age/sex brackets on Afghan refugees in Iran reported by UNHCR for end-2021 are outputs of a demographic projection model with its own uncertainty, however these counts are included without variance in the current version of the imputation model. While it would be possible to include estimates of uncertainty for the available part of the age/sex data, this would be technically more challenging and lead to more complex models. It might also be institutionally more difficult since, at least to date, most global datasets such as the one on refugee demographics have been disseminated without measures of uncertainty.

# Discussion and outlook

With the revised estimate for the proportion of children among refugees of `r round(as.numeric( t.worldregion.children |> filter(asylum_region == "WORLD") |>select(childrenMeanProp) ),0)` per cent being significantly lower than the available-data figure of `r round(as.numeric( t.worldregion.children |> filter(asylum_region == "WORLD") |>select(childrenProp) ),0)` per cent and in light of the bias in the available data shown in the descriptive analysis, we recommend use of the imputed dataset where the estimation of regional and global figures of age/sex brackets is the goal of an analysis of UNHCR's official refugee data. The imputed dataset is also suitable for analysis on the level of country of asylum or country of origin, however, in these cases particular care should be taken to analyse and communicate the uncertainty over the imputations, for example by disseminating interval- instead of point estimates. The imputed dataset contains iso3 codes for countries of origin and asylum, allowing for straightforward aggregation over custom regions and country groupings as required by an analyst. Since UNHCR's official population dataset have the same format year-on-year, fitting the same model and creating an imputed dataset from its posterior predictive distribution can be repeated for future publications of the ASR data.

@@

# References
  ---
    nocite: |
      @wpp2022, @irrs
    ...

>>>>

---
title: "Filling in data gaps on refugee children: Statistical imputation for missing count data"
author: Jan Beise, Yukun Pei (UNICEF); Sebastian Steinmueller (UNHCR)
abstract: UNHCR's annually published population dataset with the number of refugees and their age/sex composition by country of origin and country of asylum contained missing age data for 23 per cent of the global refugee populations at end-2021. Figures of the global proportion of refugees who are children disseminated to date used the available data only. We show that this approach likely leads to upwards biased estimates of the global proportion of children due to the omission of refugee groups in the Americas and Europe with lower proportions of children. We propose a multilevel Bayesian model to create a dataset with multiple imputations of age/sex counts for country pairs with missing data, resulting in standardisation of global and regional estimates by countries of origin and regions of asylum. The estimate of the proportion of children in the global refugee population derived from the imputed dataset is 42 per cent compared to 47 per cent in the available-data figure. 
format: pdf
bibliography: references.bib
execute:
  echo: false
  warning: false
  message: false
header-includes:
    \usepackage{float}
    \usepackage{setspace}
    \onehalfspacing
toc: true
lof: true
lot: true
---

---
nocite: | 
    @wpp2022, @irrs
---

@@

# Demographic data on refugees

## Reporting process and data structure

@@

## Data gaps

@@

While the imputed dataset is useful in estimating regional and global estimates of the age/sex distribution of refugees, the modelling approach is top-down and uses only geographic proximity and regional groupings as a means to improve available-data figures. It cannot substitute data collected and compiled at the national level. If the main goal is to obtain reliable estimates of the demographic characteristics of refugees on the level of country of asylum, strengthening the production process of official statistics on refugees in said country should be the primary course of action. Recommendations and implementation examples developed by countries in the framework of the Expert Group on Refugee, IDP and Statelessness Statistics (see EGRISS 2018) can provide helpful guidance to national statistical systems. Projecting populations from census, survey or administrative data on age/sex structure, fertility, mortality and displacement are another way to obtain more reliable estimates of the age/sex distribution at country level, although this requires much more effort for a single country and is therefore better suited if estimates for a specific country are required. Estimates from the imputed dataset at any geographical level should always be disseminated and communicated with uncertainty intervals or by showing the posterior predictive distribution over imputations. While care has been taken to include and test prior assumptions and hierarchical geographic structures to model uncertainty from parameters and predictions, we made a blanket assumption that all available age/sex data was measured without uncertainty and without measurement error. This is an over-simplification. For example, counts for age/sex brackets on Afghan refugees in Iran reported by UNHCR for end-2021 are outputs of a demographic projection model with its own uncertainty, however these counts are included without variance in the current version of the imputation model. While it would be possible to include estimates of uncertainty for the available part of the age/sex data, this would be technically more challenging and lead to more complex models. It might also be institutionally more difficult since, at least to date, most global datasets such as the one on refugee demographics have been disseminated without measures of uncertainty.

# Discussion

With the revised estimate for the proportion of children among refugees of `r round(as.numeric( t.worldregion.children |> filter(asylum_region == "WORLD") |>select(childrenMeanProp) ),0)` per cent being significantly lower than the available-data figure of `r round(as.numeric( t.worldregion.children |> filter(asylum_region == "WORLD") |>select(childrenProp) ),0)` per cent and in light of the bias in the available data shown in the descriptive analysis, we recommend use of the imputed dataset where the estimation of regional and global figures of age/sex brackets is the goal of an analysis of UNHCR's official refugee data. The imputed dataset is also suitable for analysis on the level of country of asylum or country of origin, however, in these cases particular care should be taken to analyse and communicate uncertainty over the imputations, for example by disseminating interval- instead of point estimates. The imputed dataset contains iso-3 codes for countries of origin and asylum, allowing for straightforward aggregation over custom regions and country groupings as required by an analyst. Since UNHCR's official population dataset has the same format year-on-year, fitting the same model and creating an imputed dataset from its posterior predictive distribution can be repeated for future publications of the ASR data.

@@

# References
